Supplementary information: Variational theory of non-relativistic quantum electrodynamics

### 2018-11-06 supp2.tex
fixed compilation

Using these derivatives, Equation (6) of the main text is generalized to:
\begin{align}
&\left(\frac{\mathbf{p}^2}{2m}+v_{ext}(\mathbf{x}) \right)\psi_i(\mathbf{x}) +  \sum\limits_{j=1}^N \int d^3x' ~ V(\mathbf{x}-\mathbf{x}')\left(\psi^*_j(\mathbf{x}')\psi_j(\mathbf{x}')\psi_i(\mathbf{x}) - \psi_j^*(\mathbf{x}')\psi_j(\mathbf{x})\psi_i(\mathbf{x}')  \right) \nonumber \\ &+ \frac{\hbar e^2}{4m\epsilon_0}\sum_n \frac{1}{\omega_n}|\mathbf{F}_n(\mathbf{x})|^2\psi_i(\mathbf{x}) + \frac{e^2\hbar^2}{8m^2\epsilon_0}\sum\limits_{n=N_{\sigma}+1}^{\infty}\sum\limits_{q=1}^{N_p} \frac{\int d^3y~\mathbf{F}^*_q(\mathbf{y})\cdot\mathbf{j}_{ni}(\mathbf{y})}{\omega_q(\omega_{in}-\lambda_q)}\left( \mathbf{F}_q(\mathbf{x})\cdot\nabla\psi_n(\mathbf{x}) + \nabla\cdot(\mathbf{F}_q(\mathbf{x})\psi_n(\mathbf{x}))\right)  = E_i\psi_i(\mathbf{x}).
\end{align}

@@

And setting the derivative with respect to the photonic frequencies to zero gives the equation:
\begin{equation}
\frac{\hbar}{4}\left(\int d^3x ~|\mathbf{F}_q(\mathbf{x})|^2 - \frac{c^2}{\omega_q^2}\mathbf{F}^*_q(\mathbf{x})\cdot\nabla\times\nabla\times\mathbf{F}_q(\mathbf{x}) \right) - \frac{e^2\hbar}{4m\epsilon_0\omega_q^2}\int d^3x~\sum\limits_{m=1}^{N_{\sigma}} |\psi_m(\mathbf{x})|^2 -\frac{e^2\hbar^2}{8m^2\epsilon_0}\sum\limits_{n=N_{\sigma}+1}^{\infty}\sum\limits_{m=1}^{N_{\sigma}}\frac{\Big|\int d^3x~\mathbf{F}_q(\mathbf{x})\cdot\mathbf{j}_{mn}(\mathbf{x})\Big|^2}{\omega_q^2(\omega_{mn}-\lambda_q)}=  0 
\end{equation}
Performing the operation $\frac{1}{\omega_q}\int d^3x \mathbf{F}_q^*(\mathbf{x})  \cdot$ on Equation (6), and adding the result to Equation (7), gives the result $\lambda_q = \omega_q$, just as before $\delta E$ was introduced. Using this key simplification, along with the definition of the spatially-dependent plasma frequency as $\omega_p^2(\mathbf{x}) = \frac{e^2}{m\epsilon_0}\sum\limits_{m=1}^{N_{\sigma}}|\psi_m(\mathbf{x})|^2$, we have that Equation (7) of the main text is generalized to:
\begin{equation}
\left( \nabla\times\nabla\times - \left(1-\frac{\omega_p^2(\mathbf{x})}{\omega_q^2} \right)\right)\mathbf{F}_q(\mathbf{x}) = -\frac{e^2\hbar}{2m^2\epsilon_0c^2}\sum\limits_{n=N_{\sigma}+1}^{\infty}\sum\limits_{m=1}^{N_{\sigma}} \frac{\int d^3y~\mathbf{F}_q(\mathbf{y})\cdot\mathbf{j}_{mn}(\mathbf{y})}{\omega_{mn}-\omega_{q}}\mathbf{j}_{nm}(\mathbf{x}).
\end{equation}
\textit{Equations (5) and (8) represent main results of this work and provide a general starting point for first-principles analysis of ground states of QED systems in the non-perturbative regime.}

@@

In this section, we provide some additional details on the one-dimensional cavity QED model considered in the main text. Given the Hamiltonian of Equation (9) in the main text describing the coupling of an emitter to a one-dimensional cavity, with the matter being described by the site model of Equations (10) and (11) of the main text, the expectation value of the Hamiltonian according to the ansatz of Equation (4) in the main text is given by
\begin{equation}
\langle \Psi | H | \Psi \rangle &= \langle \tilde{g} |H_{\text{matter}} | \tilde{g}\rangle + \frac{\hbar}{4}\int dz ~\sum\limits_{n=1}^{\infty}\left(\omega_n|F_n|^2 - \frac{c^2}{\omega_n}F_n^*\partial_z^2F_n\right) + \frac{\hbar q^2}{4m\epsilon_0\omega_n}\sum\limits_{n=1}^{\infty} \int dz~\delta(z-d)|F_n|^2
\end{equation}
In this equation, $|\tilde{g}\rangle$ is the ground state of the effective matter part of the Hamiltonian, and $H_{\mathrm{matter}}$ is the Hamiltonian of Equation (10) of the main text.  We impose constraints of matter normalization and photon mode normalization by defining a Lagrange function 
\begin{align}\label{eq:lagrange}
\mathcal{L}(|\tilde{g}\rangle, \langle \tilde{g}|,\epsilon, \{F_n, F_n^*, \omega_n, \lambda_n\}) \equiv \langle \Psi | H | \Psi \rangle - \epsilon(\langle \tilde{g}|\tilde{g}\rangle-1)-\sum\limits_{n=1}^{\infty}\frac{\hbar\lambda_n}{2}\left( \int dz~|F_n|^2-1\right).
\end{equation} 
To find the ground state, we minimize the Lagrange function with respect to the matter orbital $|\tilde{g}\rangle$ and with respect to the mode functions $F_n$. The minimization with respect to the matter leads to the trivial equation $H_{\text{matter}} |\tilde{g}\rangle = \epsilon|\tilde{g}\rangle$ which leaves the effective matter ground state as simply the ground state of $H_{\text{matter}}$. On the other hand, the minimization with respect to the photon mode functions leads to  the equation
\begin{equation}\label{eq:maxwell}
\left(\partial_z^2-\frac{\omega^2_n}{c^2}+2\frac{\omega_n\lambda_n}{c^2}-\frac{q^2 }{m\epsilon_0 c^2}\delta(z-d)\right)F_n  = 0.
\end{equation}
We may constrain the $\lambda_n$ by differentiating the Lagrange function with respect to the $\omega_n$. The equation which follows is:
\begin{equation}\label{eq:other_maxwell}
\int dz ~\left(|F_n|^2 + \frac{c^2}{\omega^2_n}F_n^*\partial_z^2F_n\right) - \frac{ q^2}{m\epsilon_0\omega^2_n} \int dz~\delta(z-d)|F_n|^2 = 0
\end{equation}
Performing $\frac{\omega_n^2}{c^2}\int dz~F_n^*$ on both sides of Equation (11), and adding this equation to Equation (12), one immediately finds that $\lambda_n = \omega_n$ and that
\begin{equation}\label{eq:final_maxwell}
\left(\partial_z^2+\frac{\omega^2_n}{c^2}-\frac{q^2 }{m\epsilon_0 c^2}\delta(z-d)\right)F_n  = 0.
\end{equation}

This is an ordinary second-order differential equation with the conditions that $F_n$ is continuous at $d$ and that its derivative is discontinuous according to 
\begin{equation}\label{eq:boundary_condition}
\partial_zF_n\Big|_{z=d^+}-\partial_zF_n\Big|_{z=d^-} = \frac{q^2}{m\epsilon_0 c^2}F_n(d),
\end{equation}
in addition to the usual condition of the modes vanishing at the cavity walls $z=0$ and $z=L$. It can be shown that the solution to Equation (13) satisfying such boundary conditions is:
\begin{align}\label{eq:field_mode}
&\theta (z-d) \left(\frac{\sin\left(\frac{\omega_nL}{c}\right)\sin\left(\frac{\omega_nd}{c}\right)\cos\left(\frac{\omega_nz}{c}\right)}{\sin\left(\frac{\omega_n(L-d)}{c}\right)}-\right) \nonumber \\ 
-&\theta (z-d) \left(\frac{\cos\left(\frac{\omega_nL}{c}\right)\sin\left(\frac{\omega_nd}{c}\right)\sin\left(\frac{\omega_nz}{c}\right)}{\sin\left(\frac{\omega_n(L-d)}{c}\right)}\right) \nonumber \\ 
+&\theta (d-z) \sin\left(\frac{\omega_n z}{c} \right)
\end{align}
provided that the auxiliary condition
\begin{equation}\label{eq:mode_transcendental}
\cot\left(\frac{\omega_n}{c}d \right)+\cot\left(\frac{\omega_n}{c}(L-d) \right) = -\frac{q^2}{m\epsilon_0\omega_nc}
\end{equation}
is met. To ensure that the modes are normalized according to the constraint, we have that the solutions of Equation (15) must be multiplied by a normalization factor $N_n$ given by
\begin{equation}\label{eq:mode_normalization}
N_n = 2\sqrt{\frac{1}{\frac{c}{\omega_n}\left(\frac{\omega_nL}{c}-\sin\left(\frac{\omega_nL}{c}\right) \right)\left(1+\frac{\sin^2\left(\frac{\omega_nd}{c}\right)}{\sin^2\left(\frac{\omega_n(L-d)}{c}\right)} \right)}}.
\end{equation}
The condition of Equation \ref{eq:mode_transcendental} determines the resonance frequencies of the photon quasiparticle modes. \Jadd{(put eq.12-14 to SI?)}

@@

Here, we note the parameters used in Fig. 2.
\begin{enumerate}
\item{The hopping matrix elements $t$ were taken to be 0.25 eV for the two-, three-, and four-level systems. Meanwhile, the on-site energies were taken to be equal on all sites in the two-, three-, and four-level systems.}
\item{The cavity length was taken to be 1 micron.}
\item{The area of the cavity in the transverse direction was taken to be 100 nm$^2$.}
\item{The maximum number of cavity modes retained in the calculations was 100. Our results were converged with respect to the number of cavity modes.}
\item{In the numerical diagonalization results (red lines of Fig. 2 of the main text), the Fock space was truncated such that the number of photons retained was no more than four. For the largest couplings plotted in Fig. 2, this was sufficient. But for higher couplings, more photons in the numerical diagonalization are needed. For four photons and 100 cavity modes coupled to a four-level system, the dimension of the Hilbert space is $4\times 10^8$. }
\item{The largest couplings plotted in Fig. 2 correspond to either a single emitter with a charge of $200e$, or an ensemble of emitters (as is the case in many ultra-strong coupling experiments) in which there are 40,000 emitters in the cavity. }
\end{enumerate}

>>>>

Using these derivatives, Equation (6) of the main text is generalized to:
\begin{align}
&\left(\frac{\mathbf{p}^2}{2m}+v_{ext}(\mathbf{x}) \right)\psi_i(\mathbf{x}) +  \sum\limits_{j=1}^N \int d^3x' ~ V(\mathbf{x}-\mathbf{x}')\left(\psi^*_j(\mathbf{x}')\psi_j(\mathbf{x}')\psi_i(\mathbf{x}) - \psi_j^*(\mathbf{x}')\psi_j(\mathbf{x})\psi_i(\mathbf{x}')  \right) \nonumber \\ &+ \frac{\hbar e^2}{4m\epsilon_0}\sum_n \frac{1}{\omega_n}|\mathbf{F}_n(\mathbf{x})|^2\psi_i(\mathbf{x}) + \frac{e^2\hbar^2}{8m^2\epsilon_0}\sum\limits_{n=N_{\sigma}+1}^{\infty}\sum\limits_{q=1}^{N_p} \frac{\int d^3y~\mathbf{F}^*_q(\mathbf{y})\cdot\mathbf{j}_{ni}(\mathbf{y})}{\omega_q(\omega_{in}-\lambda_q)}\left( \mathbf{F}_q(\mathbf{x})\cdot\nabla\psi_n(\mathbf{x}) + \nabla\cdot(\mathbf{F}_q(\mathbf{x})\psi_n(\mathbf{x}))\right)  \nonumber \\ &= E_i\psi_i(\mathbf{x}).
\end{align}

@@

And setting the derivative with respect to the photonic frequencies to zero gives the equation:
\begin{align}
&\frac{\hbar}{4}\left(\int d^3x ~|\mathbf{F}_q(\mathbf{x})|^2 - \frac{c^2}{\omega_q^2}\mathbf{F}^*_q(\mathbf{x})\cdot\nabla\times\nabla\times\mathbf{F}_q(\mathbf{x}) \right) - \frac{e^2\hbar}{4m\epsilon_0\omega_q^2}\int d^3x~\sum\limits_{m=1}^{N_{\sigma}} |\psi_m(\mathbf{x})|^2 \nonumber \\ &-\frac{e^2\hbar^2}{8m^2\epsilon_0}\sum\limits_{n=N_{\sigma}+1}^{\infty}\sum\limits_{m=1}^{N_{\sigma}}\frac{\Big|\int d^3x~\mathbf{F}_q(\mathbf{x})\cdot\mathbf{j}_{mn}(\mathbf{x})\Big|^2}{\omega_q^2(\omega_{mn}-\lambda_q)}=  0 
\end{align}
Performing the operation $\frac{1}{\omega_q}\int d^3x \mathbf{F}_q^*(\mathbf{x})  \cdot$ on Equation (6), and adding the result to Equation (7), gives the result $\lambda_q = \omega_q$, just as before $\delta E$ was introduced. Using this key simplification, along with the definition of the spatially-dependent plasma frequency as $\omega_p^2(\mathbf{x}) = \frac{e^2}{m\epsilon_0}\sum\limits_{m=1}^{N_{\sigma}}|\psi_m(\mathbf{x})|^2$, we have that Equation (7) of the main text is generalized to:
\begin{equation}
\left( \nabla\times\nabla\times - \left(1-\frac{\omega_p^2(\mathbf{x})}{\omega_q^2} \right)\right)\mathbf{F}_q(\mathbf{x}) = -\frac{e^2\hbar}{2m^2\epsilon_0c^2}\sum\limits_{n=N_{\sigma}+1}^{\infty}\sum\limits_{m=1}^{N_{\sigma}} \frac{\int d^3y~\mathbf{F}_q(\mathbf{y})\cdot\mathbf{j}_{mn}(\mathbf{y})}{\omega_{mn}-\omega_{q}}\mathbf{j}_{nm}(\mathbf{x}).
\end{equation}
\textit{Equations (5) and (8) represent main results of this work and provide a general starting point for first-principles analysis of ground states of QED systems in the non-perturbative regime.}

@@

In this section, we provide some additional details on the one-dimensional cavity QED model considered in the main text. Given the Hamiltonian of Equation (9) in the main text describing the coupling of an emitter to a one-dimensional cavity, with the matter being described by the site model of Equations (10) and (11) of the main text, the expectation value of the Hamiltonian according to the ansatz of Equation (4) in the main text is given by
\begin{equation}
\langle \Psi | H | \Psi \rangle = \langle \tilde{g} |H_{\text{matter}} | \tilde{g}\rangle + \frac{\hbar}{4}\int dz ~\sum\limits_{n=1}^{\infty}\left(\omega_n|F_n|^2 - \frac{c^2}{\omega_n}F_n^*\partial_z^2F_n\right) + \frac{\hbar q^2}{4m\epsilon_0\omega_n}\sum\limits_{n=1}^{\infty} \int dz~\delta(z-d)|F_n|^2
\end{equation}
In this equation, $|\tilde{g}\rangle$ is the ground state of the effective matter part of the Hamiltonian, and $H_{\mathrm{matter}}$ is the Hamiltonian of Equation (10) of the main text.  We impose constraints of matter normalization and photon mode normalization by defining a Lagrange function 
\begin{equation}\label{eq:lagrange}
\mathcal{L}(|\tilde{g}\rangle, \langle \tilde{g}|,\epsilon, \{F_n, F_n^*, \omega_n, \lambda_n\}) \equiv \langle \Psi | H | \Psi \rangle - \epsilon(\langle \tilde{g}|\tilde{g}\rangle-1)-\sum\limits_{n=1}^{\infty}\frac{\hbar\lambda_n}{2}\left( \int dz~|F_n|^2-1\right).
\end{equation} 
To find the ground state, we minimize the Lagrange function with respect to the matter orbital $|\tilde{g}\rangle$ and with respect to the mode functions $F_n$. The minimization with respect to the matter leads to the trivial equation $H_{\text{matter}} |\tilde{g}\rangle = \epsilon|\tilde{g}\rangle$ which leaves the effective matter ground state as simply the ground state of $H_{\text{matter}}$. On the other hand, the minimization with respect to the photon mode functions leads to  the equation
\begin{equation}\label{eq:maxwell}
\left(\partial_z^2-\frac{\omega^2_n}{c^2}+2\frac{\omega_n\lambda_n}{c^2}-\frac{q^2 }{m\epsilon_0 c^2}\delta(z-d)\right)F_n  = 0.
\end{equation}
We may constrain the $\lambda_n$ by differentiating the Lagrange function with respect to the $\omega_n$. The equation which follows is:
\begin{equation}\label{eq:other_maxwell}
\int dz ~\left(|F_n|^2 + \frac{c^2}{\omega^2_n}F_n^*\partial_z^2F_n\right) - \frac{ q^2}{m\epsilon_0\omega^2_n} \int dz~\delta(z-d)|F_n|^2 = 0
\end{equation}
Performing $\frac{\omega_n^2}{c^2}\int dz~F_n^*$ on both sides of Equation (11), and adding this equation to Equation (12), one immediately finds that $\lambda_n = \omega_n$ and that
\begin{equation}\label{eq:final_maxwell}
\left(\partial_z^2+\frac{\omega^2_n}{c^2}-\frac{q^2 }{m\epsilon_0 c^2}\delta(z-d)\right)F_n  = 0.
\end{equation}

This is an ordinary second-order differential equation with the conditions that $F_n$ is continuous at $d$ and that its derivative is discontinuous according to 
\begin{equation}\label{eq:boundary_condition}
\partial_zF_n\Big|_{z=d^+}-\partial_zF_n\Big|_{z=d^-} = \frac{q^2}{m\epsilon_0 c^2}F_n(d),
\end{equation}
in addition to the usual condition of the modes vanishing at the cavity walls $z=0$ and $z=L$. It can be shown that the solution to Equation (13) satisfying such boundary conditions is:
\begin{align}\label{eq:field_mode}
&\theta (z-d) \left(\frac{\sin\left(\frac{\omega_nL}{c}\right)\sin\left(\frac{\omega_nd}{c}\right)\cos\left(\frac{\omega_nz}{c}\right)}{\sin\left(\frac{\omega_n(L-d)}{c}\right)}-\right) \nonumber \\ 
-&\theta (z-d) \left(\frac{\cos\left(\frac{\omega_nL}{c}\right)\sin\left(\frac{\omega_nd}{c}\right)\sin\left(\frac{\omega_nz}{c}\right)}{\sin\left(\frac{\omega_n(L-d)}{c}\right)}\right) \nonumber \\ 
+&\theta (d-z) \sin\left(\frac{\omega_n z}{c} \right)
\end{align}
provided that the auxiliary condition
\begin{equation}\label{eq:mode_transcendental}
\cot\left(\frac{\omega_n}{c}d \right)+\cot\left(\frac{\omega_n}{c}(L-d) \right) = -\frac{q^2}{m\epsilon_0\omega_nc}
\end{equation}
is met. To ensure that the modes are normalized according to the constraint, we have that the solutions of Equation (15) must be multiplied by a normalization factor $N_n$ given by
\begin{equation}\label{eq:mode_normalization}
N_n = 2\sqrt{\frac{1}{\frac{c}{\omega_n}\left(\frac{\omega_nL}{c}-\sin\left(\frac{\omega_nL}{c}\right) \right)\left(1+\frac{\sin^2\left(\frac{\omega_nd}{c}\right)}{\sin^2\left(\frac{\omega_n(L-d)}{c}\right)} \right)}}.
\end{equation}
The condition of Equation \ref{eq:mode_transcendental} determines the resonance frequencies of the photon quasiparticle modes.

@@

Here, we note the parameters used in Fig. 2.
\begin{enumerate}
\item{The hopping matrix elements $t$ were taken to be 0.25 eV for the two-, three-, and four-level systems. Meanwhile, the on-site energies were taken to be equal on all sites in the two-, three-, and four-level systems.}
\item{The cavity length was taken to be 1 micron.}
\item{The area of the cavity in the transverse direction was taken to be 100 nm$^2$.}
\item{The maximum number of cavity modes retained in the calculations was 100. Our results were converged with respect to the number of cavity modes.}
\item{In the numerical diagonalization results (red lines of Fig. 2 of the main text), the Fock space was truncated such that the number of photons retained was no more than four. For the largest couplings plotted in Fig. 2, this was sufficient. But for higher couplings, more photons in the numerical diagonalization are needed. For four photons and 100 cavity modes coupled to a four-level system, the dimension of the Hilbert space is $4\times 10^8$. }
\item{The largest couplings plotted in Fig. 2 correspond to either a single emitter with a charge of $200e$, or an ensemble of emitters (as is the case in many ultra-strong coupling experiments) in which there are 40,000 emitters in the cavity. The largest couplings correspond to rather extreme coupling parameters and are shown mostly to demonstrate that our ansatz is quite accurate even in regimes of extremely high coupling.}
\end{enumerate}
